Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-28 paper/perfmod.tex
do not forget tuminaro

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012}.  However, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015} models use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

@@

For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.  By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.  Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the number of preconditioned Newton-Krylov iterations \citep{Bueler2021} grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scaling.m
The algebraic-multigrid based solver in \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.  A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.

@@

FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems; large constants are acceptable if $\beta,\gamma$ are less than 2, for example.  Specifically $\gamma < 1$ might be regarded as the goal of the next decades of research on numerical ice sheet models using Stokes dynamics.  However, this aspiration is long-term because work is barely started on it.

>>>>

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

The assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.  This and other mesh/solver details cannot be pursued here.

@@

For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.

By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.  Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.  \cite[Table 7.5]{Tuminaroetal2016} reports total algebraic-multigrid-preconditioned iterations growing as $\alpha \approx 0.05$ in a higher-order velocity solver for the Greenland ice sheet.  (Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.)
% see data/scaling.m
An algebraic-multigrid based solver by \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.  A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012,Tumarinoetal2016} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.

@@

The real promise of near-optimal methods is revealed when implicit solvers are considered, ones which simultanously satisfy the mass and momentum equations at each time step.  If a multigrid method can be designed which yields $\gamma\approx 0$ for a Stokes model, for example, then numerical ice sheet modeling tools will be qualitatively faster for many tasks, and a large constant would be acceptable.  Specifically, and more modestly, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method is a goal for the next decades of research on numerical ice sheet models.  Such a tool would scale at $O(\Delta x^{-3})$, better than current explicit SIA models, avoiding shallow approximations while computing time steps with a frequency determined only by the scientific need to resolve climate coupling.  The same goal applies to higher-order, or other membrane-stress-resolving, solvers.   These aspirations are admittedly long-term because, to the author's knowledge, no progress has yet been made on such implicit designs.
